---
title: "**Exam Physiology: Wearable Data**"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document
---

This activity will walk through interpretation of data from wearable devices intended to measure stress in humans. You will work with real data from a study where researchers measured stress in students taking exams.

## Part 1

In Part 1, we'll go over some background context for this activity. Then, we'll get set up on Posit Cloud for the next steps.

### Background

First, let's read the "plain language" summary from a publication entitled "A Wearable Exam Stress Dataset for Predicting Grades using Physiological Signals" [@amin2022wearable].

> Most stress research happens in labs, not real life. This means there aren't many datasets showing how people's bodies respond to actual stressful situations. We measured stress in 10 students during three real exams using wearable sensors that tracked skin conductance (how much their skin conducted electricity—a sign of stress) and skin temperature. We then looked at patterns in the data to see if we could predict whether students would get high or low grades based on their stress levels. Our prediction method was 70-80% accurate at identifying high versus low performers. We also found general patterns showing how stress levels changed throughout the exams. This shows that wearable devices could potentially be used to monitor stress and predict performance during real-world stressful situations. We're making this dataset publicly available so other researchers can use it to tackle the challenges of measuring stress outside the lab.

As the summary suggests, lab settings do not always reflect reality. Most stress research uses controlled laboratory settings with standardized stressors like electric shocks, loud noises, or disturbing images. These methods allow researchers to isolate specific physiological responses and build reliable datasets. This approach has helped us understand how the autonomic nervous system responds to stress. Howevever, the challenge is that Real cognitive stressors like academic exams involve anticipatory anxiety, sustained pressure, and outcomes that actually matter to participants. The physiological responses may differ substantially. This "real world" research is important for validating findings from the lab.

**Electrodermal activity**, also known as skin conductance or EDA, measures how electrical properties of the skin change over time. When you are stressed or anxious, your sympathetic nervous system activates your sweat glands, which changes how well your skin conducts electricity. More stress generally means greater EDA. This makes EDA a good indicator of psychological and physiological stress.

::: callout-note
## Check Your Knowledge

1. What variables were the researchers trying to measure?

2.  True or False: High Electrodermal activity (EDA) means the person being measured might be stressed.
:::

### Start Posit Cloud

In the next steps, we will look at the data from the publication described above. We will be working in a programming language called R that is widely used for analysis and making graphs/plots. We'll use the version of R available online on Posit Cloud.

Go to <https://posit.cloud/plans/free> and follow the steps to sign up. You can also Log In using the button on the top right if you have used Posit Cloud in the past.

<!-- callout about what cloud computing is ? -->

```{r, fig.align='center', echo = FALSE, fig.alt= "TO ADD", out.width = '90%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1dhP4CeUIJAqy3z7mKxuFzk8b8LhLrPBPHb1bpYpbk3c/edit?slide=id.g1965a5f7f0a_0_44#slide=id.g1965a5f7f0a_0_44")
```

Once logged in, select "New Project" and "New RStudio Project". It will take a few seconds to deploy and load.

```{r, fig.align='center', echo = FALSE, fig.alt= "TO ADD", out.width = '90%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1dhP4CeUIJAqy3z7mKxuFzk8b8LhLrPBPHb1bpYpbk3c/edit?slide=id.g357ffc10808_0_3#slide=id.g357ffc10808_0_3")
```

Rename your project "Exam Physiology" by clicking on "Untitled Project".

```{r, fig.align='center', echo = FALSE, fig.alt= "TO ADD", out.width = '90%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1dhP4CeUIJAqy3z7mKxuFzk8b8LhLrPBPHb1bpYpbk3c/edit?slide=id.g357ffc10808_0_12#slide=id.g357ffc10808_0_12")
```

### Package Install and Load

We will need to install the `tidyverse` package for this activity.

Type the following into your console and press return to run the code. You will see a lot of text, but usually that's a good thing!

```{r message=FALSE}
install.packages("tidyverse")
```

<!-- ```{r, fig.align='center', echo = FALSE, fig.alt= "TO ADD", out.width = '90%'} -->
<!-- ottrpal::include_slide("https://docs.google.com/presentation/d/1dhP4CeUIJAqy3z7mKxuFzk8b8LhLrPBPHb1bpYpbk3c/edit?slide=id.g357ffc10808_0_30#slide=id.g357ffc10808_0_30") -->
<!-- ``` -->

Next, we will load the package so it's ready to use:

```{r message=FALSE}
library(tidyverse)
```

<!-- ```{r, fig.align='center', echo = FALSE, fig.alt= "TO ADD", out.width = '90%'} -->
<!-- ottrpal::include_slide("https://docs.google.com/presentation/d/1dhP4CeUIJAqy3z7mKxuFzk8b8LhLrPBPHb1bpYpbk3c/edit?slide=id.g357ffc10808_0_61#slide=id.g357ffc10808_0_61") -->
<!-- ``` -->

::: callout-tip
## What are R Packages?

Packages are collections of R code, data, and documentation that extend the base functionality of R. Think of them like "expansion packs" on top of your basic R software.

Packages are developed by the R community and made available through repositories like CRAN (Comprehensive R Archive Network), Bioconductor, and GitHub. They are especially useful if you want to do a specialized kind of analysis, such as genomic analysis!

We use the `library` command to load and attach packages to the R environment. This means links the package you downloaded to your current session of R.
:::

### Get Physiology Data

We'll read in the data from this study the GEMs website. Run the following code in your console:

```{r}
phys_data <- read.csv("https://genomicseducation.org/data/phys_data.csv")
head(phys_data)
```

If everything went correctly, you should see both an object called `phys_data` in the Environment Pane (top right) and the data printed to the console.

<!-- ```{r, fig.align='center', echo = FALSE, fig.alt= "TO ADD", out.width = '90%'} -->
<!-- ottrpal::include_slide("https://docs.google.com/presentation/d/1dhP4CeUIJAqy3z7mKxuFzk8b8LhLrPBPHb1bpYpbk3c/edit?slide=id.g357ffc10808_0_47#slide=id.g357ffc10808_0_47") -->
<!-- ``` -->

::: callout-tip
## What is an R object?

An object is something stored in a programming language, such as data. Think of it like a "noun".

In R, we assign objects using the `<-` symbol, with the name of your choice on the left, and the contents of the object on the right.

Objects will show up in your Environment Pane. It's a good idea to check the Environment Pane regularly.
:::

Enter the following code in the console to view the data in tabular format:

```{r eval=F}
View(phys_data)
```

<!-- ```{r, fig.align='center', echo = FALSE, fig.alt= "TO ADD", out.width = '90%'} -->
<!-- ottrpal::include_slide("https://docs.google.com/presentation/d/1dhP4CeUIJAqy3z7mKxuFzk8b8LhLrPBPHb1bpYpbk3c/edit?slide=id.g357ffc10808_0_74#slide=id.g357ffc10808_0_74") -->
<!-- ``` -->

### Data Dictionaries

Take a look at the column names in the Viewer you just opened. Compare these to the data dictionary below to interpret what the columns represent.

::: callout-tip
## Pedigree Data Dictionary

**timestamp**: Time the measurement was taken. The measurement comes from a wearable device worn by a student.\
**eda**: Measurement from the electrodermal activity sensor.\
**hr**: Heart rate measurement, in beats per minute.\
**temp**: Skin temperature measurement, in Celcius.\
**test**: Which test/exam was being taken.\
**studentID**: Student being measured.\
:::

### Part 1 Questions

::: callout-note
## Check Your Knowledge

1.  Why might we want to install add-on packages in R?

2.  How can we confirm that our data loaded correctly? Take a screenshot and point to your evidence.

3.  Looking at the `timestamp` column, how often are measurements being collected?

4.  How many rows of data, or observations, does this data have? Hint: look at the Environment Pane (top right) and see how many `obs.` you have.
:::

## Part 2

Now, we'll run some R code to better understand our data.

### Counting Tests and Students

First, let's see how many different tests/exams are represented, and how many observations we have of each.

```{r}
count(phys_data, test)
```

Looks like we have three tests: Midterm 1, Midterm 2, and the Final exam.

Now let's look at how many different students are represented, and how many observations we have of each.

```{r}
count(phys_data, studentID)
```

Looks like we have 10 students who were measured. The number of observations vary a bit because some students might have had their sensors turned on a bit longer than others.

### Part 2a Questions

::: callout-note
## Check Your Knowledge

1.  Run `count(phys_data, test, studentID)`. How many observations does student `S1` have for Midterm 1? Take a screenshot showing this number.

2.  Which student & test combination has the lowest number of observations?
:::

### Sorting in Viewer

In the data Viewer (top left), click on the variable/column names to sort them. For example, If we click `timestamp`, we can see the earliest timestamp is `2018-10-13 12:55:31`. We can click it again to see the latest timestamp is `2018-10-13 12:55:31`.

<!-- ```{r, fig.align='center', echo = FALSE, fig.alt= "TO ADD", out.width = '90%'} -->
<!-- ottrpal::include_slide("https://docs.google.com/presentation/d/1dhP4CeUIJAqy3z7mKxuFzk8b8LhLrPBPHb1bpYpbk3c/edit?slide=id.g357ffc10808_0_30#slide=id.g357ffc10808_0_30") -->
<!-- ``` -->

<!-- ```{r, fig.align='center', echo = FALSE, fig.alt= "TO ADD", out.width = '90%'} -->
<!-- ottrpal::include_slide("https://docs.google.com/presentation/d/1dhP4CeUIJAqy3z7mKxuFzk8b8LhLrPBPHb1bpYpbk3c/edit?slide=id.g357ffc10808_0_30#slide=id.g357ffc10808_0_30") -->
<!-- ``` -->

### Part 2b Questions

::: callout-note
## Check Your Knowledge

1.  Sort the `hr` column. What is the greatest heart rate observed? Which student was this? Take a screenshot of your evidence.

2.  Which two students had the greatest `eda` values? Take a screenshot of your evidence.
:::
