---
title: "**Family Markers: Using Multiply-Affected Families to Identify Risk Genes**"
output:
  html_document
---

## Materials

You will need an internet connection for this activity as written. Students can use a local installation of R or RStudio if the data is downloaded ahead of time.

Download this module as:

-   Quarto (qmd)
-   Rmarkdown (Rmd)
-   Word (docx)

Download the R student activity as:

-   Quarto (qmd)
-   Rmarkdown (Rmd)
-   Word (docx)

Borrow images from [Google Slides](https://docs.google.com/presentation/d/1fy3FwkSLETnRUdf8huddq2n4qcb6eOyzwVg05dbudII/edit?usp=sharing).

## Learning Objectives

Coming soon!

## Activity 1 - Literature Review

Coming soon!

### Questions

## Activity 2 - Pedigree Construction in R

In this activity. you will ..

### Start Posit Cloud

We will use Posit Cloud for this activity. Go to <https://posit.cloud/plans/free> and follow the steps to sign up. You can also Log In using the button on the top right if you have used Posit Cloud in the past.

<!-- Callout about what cloud computing is -->

```{r, fig.align='center', echo = FALSE, fig.alt= "TO ADD", out.width = '90%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fy3FwkSLETnRUdf8huddq2n4qcb6eOyzwVg05dbudII/edit?slide=id.g1965a5f7f0a_0_44#slide=id.g1965a5f7f0a_0_44")
```

Once logged in, select "New Project" and "New RStudio Project". It will take a few seconds to deploy and load.

```{r, fig.align='center', echo = FALSE, fig.alt= "TO ADD", out.width = '90%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fy3FwkSLETnRUdf8huddq2n4qcb6eOyzwVg05dbudII/edit?slide=id.g357ffc10808_0_3#slide=id.g357ffc10808_0_3")
```

Rename your project "Family Markers" by clicking on "Untitled Project".

```{r, fig.align='center', echo = FALSE, fig.alt= "TO ADD", out.width = '90%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fy3FwkSLETnRUdf8huddq2n4qcb6eOyzwVg05dbudII/edit?slide=id.g357ffc10808_0_12#slide=id.g357ffc10808_0_12")
```

### Package Install and Load

We will need to install the `kinship2` package for this activity.

Type the following into your console and press return to run the code. You will see a lot of red text, but usually that's a good thing!

```{r eval=F}
install.packages("kinship2")
```

```{r, fig.align='center', echo = FALSE, fig.alt= "TO ADD", out.width = '90%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fy3FwkSLETnRUdf8huddq2n4qcb6eOyzwVg05dbudII/edit?slide=id.g357ffc10808_0_30#slide=id.g357ffc10808_0_30")
```

Next, we will load the package so it's ready to use:

```{r eval=F}
library(kinship2)
```

```{r, fig.align='center', echo = FALSE, fig.alt= "TO ADD", out.width = '90%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fy3FwkSLETnRUdf8huddq2n4qcb6eOyzwVg05dbudII/edit?slide=id.g357ffc10808_0_61#slide=id.g357ffc10808_0_61")
```

::: {.callout-tip}
## What are R Packages?

Packages are collections of R code, data, and documentation that extend the base functionality of R. Think of them like "expansion packs" on top of your basic R software.

Packages are developed by the R community and made available through repositories like CRAN (Comprehensive R Archive Network), Bioconductor, and GitHub. They are especially useful if you want to do a specialized kind of analysis, such as genomic analysis!

We use the `library` command to load and attach packages to the R environment. This means links the package you downloaded to your current session of R.
:::

### Get Pedigree Data

We'll read in some data from the GEMs website. Run the following code in your console:

```{r}
ped_data <- read.csv("https://genomicseducation.org/data/pedigree_data.csv")
ped_data
```

If everything went correctly, you should see both an object called `ped_data` in the Environment Pane (top right) and the data printed to the console.

```{r, fig.align='center', echo = FALSE, fig.alt= "TO ADD", out.width = '90%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fy3FwkSLETnRUdf8huddq2n4qcb6eOyzwVg05dbudII/edit?slide=id.g357ffc10808_0_47#slide=id.g357ffc10808_0_47")
```

::: {.callout-tip}
## What is an R object?

An object is something stored in a programming language, such as data. Think of it like a "noun". 

In R, we assign objects using the `<-` symbol, with the name of your choice on the left, and the contents of the object on the right. 

Objects will show up in your Environment Pane. It's a good idea to check the Environment Pane regularly.
:::

Enter the following code in the console to view the data in tabular format:

```{r eval=F}
View(ped_data)
```

```{r, fig.align='center', echo = FALSE, fig.alt= "TO ADD", out.width = '90%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fy3FwkSLETnRUdf8huddq2n4qcb6eOyzwVg05dbudII/edit?slide=id.g357ffc10808_0_74#slide=id.g357ffc10808_0_74")
```

### Questions

::: {.callout-note}
## Check Your Knowledge

1. Why might we want to install add-on packages in R?

2. How can we confirm that our data loaded correctly?

3. Assume the column "bipolar" is 1=yes, 0=no, and NA=missing. Assume the column "depression" is the same. Based on this information, what do you think the column "diagnosis_available" represents?
:::
