---
title: "**Family Markers: Using Multiply-Affected Families to Identify Risk Genes**"
output:
  html_document
---

## Materials

Download this module as:

-   [Quarto (qmd)](https://github.com/fhdsl/GEMs/blob/main/module/family_markers.qmd)
-   Word (docx)

Download the R student activity as:

-   Quarto (qmd)
-   Word (docx)

## Learning Objectives

Coming soon!

## Activity 1 - Literature Review

Coming soon!

### Questions

## Activity 2 - Pedigree Construction in R

In this activity. you will ..

### Start Posit Cloud

We will use Posit Cloud for this activity. Go to [https://posit.cloud/plans/free](https://posit.cloud/plans/free) and follow the steps to sign up. You can also Log In using the button on the top right if you have used Posit Cloud in the past.

<!-- Callout about what cloud computing is -->

```{r, fig.align='center', echo = FALSE, fig.alt= "TO ADD", out.width = '90%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fy3FwkSLETnRUdf8huddq2n4qcb6eOyzwVg05dbudII/edit?slide=id.g1965a5f7f0a_0_44#slide=id.g1965a5f7f0a_0_44")
```

Once logged in, select "New Project" and "New RStudio Project". It will take a few seconds to deploy and load.

```{r, fig.align='center', echo = FALSE, fig.alt= "TO ADD", out.width = '90%'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fy3FwkSLETnRUdf8huddq2n4qcb6eOyzwVg05dbudII/edit?slide=id.g357ffc10808_0_3#slide=id.g357ffc10808_0_3")
```

### Get the data

### Questions
