---
title: "**Gut-brain axis: Autism Spectrum Disorder: Student Guide**"
format:
    html: default
---

```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

This module introduces students to differential expression analyses using the R programming language.

This activity will walk through differential expression analysis and use bioinformatics tools (R) to understand how gut bacteria can influence the expression of genes in the brain (the gut-brain axis). You will work with real data from a mouse RNA-seq experiment to understand how researchers explore how human disorders might be impacted by gut-brain axis using model organisms.

## Part 1

In Part 1, we'll go over some background context for this activity. Then, we'll get set up on Posit Cloud for the next steps.

The data for this activity includes gene expression data from two different brain regions (striatum and prefrontal cortex) in mice. Mice in this experiment received fecal transplants. Some mice received transplants from humans who have been diagnosed with Autism Spectrum Disorder (ASD), while other mice received transplants from humans who did not have any diagnosis (control). This allowed researchers to control the composition of the gut microbiome in each mouse. Researchers then bred the mice and looked for differences in gene expression in the brains of the offspring.

### Background
_Autism spectrum disorder (ASD)_ is a neurological disorder that affects behavioral and social interactions, among other things. Although ASD can be diagnosed at any age, it’s considered a neurodevelopmental disorder because symptoms usually show up within the first two years of life. Individuals diagnosed with ASD can experience a wide range of symptoms, including differences in social behaviors and communication styles, as well as intellectual disabilities and physical issues like sensory sensitivities or gastrointestinal problems.

The _striatum_ is part of the brain involved in motor control and cognitive tasks like reward processing, decision-making, and social interactions (often called executive functions). It lies deep within the center of the brain and is composed of both gray matter (which can be thought of as the “processing” part of brain tissue) and white matter (which is the brain structure involved in transporting messages); the combination of the gray and white matter give this region of the brain a striped appearance, resulting in the name “striatum”. The striatum is involved in both reflexive movement - that is, involuntary movement that happens as an immediate response to a stimulus - and slower, planned movement like walking. In Parkinson’s disease, some patients experience degeneration of parts of the striatum, resulting in spastic, uncontrollable movement.

The _prefrontal cortex_ is the part of the brain that is primarily in charge of decision making, reasoning, personality, maintaining social appropriateness, and other complex behaviors that fall under the umbrella of executive functions. This can include planning, self-control, and working towards long-term goals. The prefrontal cortex is located in the very front of the brain, just behind your forehead. One of the most famous brain injury patients was Phineas Gage, a railroad worker who survived an iron rod through his forehead. His prefrontal cortex was destroyed in this accident, and doctors noted huge behavioral and personality changes. You can read more about his case [here](https://pmc.ncbi.nlm.nih.gov/articles/PMC1114479/).

#### The original study

The original study, _Gut Microbiota from Human Autism Spectrum Disorder Induces Behavioral Deficits in Mice_, was published in 2019. Gut microbiota are known to be different between individuals with ASD and individuals who are considered typically-developing. Additionally, some individuals with ASD also experience gastrointestinal symptoms, and their gut microbiota show the greatest difference when compared to the gut microbiota of typically-developing individuals. Some researchers have proposed that gut bacteria can influence some of the symptoms of ASD. The relationship between the intestinal microbiome and the development and function of the human brain is known as the _gut-brain axis._

In this study, researchers explored whether they could induce ASD-like behaviors in mice by changing their gut microbiome. They transplanted stool from either humans with ASD or controls into germ-free mice and discovered that colonization with gut microbiota was enough to induce ASD-like behaviors in the mice. They also let the mice breed and collected gene expression data from the brains of their offspring to explore whether changing the gut microbiota could result in changed gene expression. In particular, they discovered that the offspring of mice who received stool from ASD donors showed different gene splicing and expression profiles of certain ASD-relevant genes.

:::{.callout-note}
It is important to note that researchers are **not** suggesting that ASD is entirely induced by gut bacteria. There is a strong genetic component to ASD. Scientists have known for years that there are **both genetic and environmental components** to the development or severity of **some** ASD symptoms. This research explores one possible environmental component.
:::

You can read the original research paper [here](https://pmc.ncbi.nlm.nih.gov/articles/PMC6993574/pdf/nihms-1528786.pdf).

#### The mouse as a model organism

The mouse is the most commonly-used model organism in laboratory work. In fact, mice and rats make up 95% of the lab animal population, and more than 80% of the research that has been awarded the Nobel Prize for Medicine was done at least in part with mouse models (https://www.cshl.edu/of-mice-and-model-organisms/, https://fbresearch.org/medical-advances/nobel-prizes). 

So what makes mice such good model organisms for biomedical research? Well, first, they’re economical and relatively easy to keep. Since mice are small, they don’t require a huge amount of space or food. They also have fast reproductive cycles, so researchers can study multiple generations within only a few years. Most importantly, though, mice and humans are both mammals and have about 85% of their protein-coding genome in common. As a result, mouse physiology is quite similar to human physiology. The mouse circulatory, reproductive, digestive, hormonal, and nervous systems are frequently used as models to study how humans grow, age, and develop chronic diseases. They are particularly important model organisms for cancer research and neuroscience. 

You can find additional information about how the mouse is used in research [here!](https://www.google.com/url?q=https://www.informatics.jax.org/mgihome/other/homepage_IntroMouse.shtml)

### Start Posit Cloud

In the next steps, we will explore the data from the publication described above.

We will use Posit Cloud for this activity. Go to <https://posit.cloud/plans/free> and follow the steps to sign up. You can also Log In using the button on the top right if you have used Posit Cloud in the past.

<!-- callout about what cloud computing is ? -->

TODO: Add figure

Once logged in, select "New Project" and "New RStudio Project". It will take a few seconds to deploy and load.

TODO: Add figure

Rename your project "Differential Expression ASD" by clicking on "Untitled Project".

TODO: Add figure

### Package Install and Load

We will need to install the `tidyverse` package for this activity.

Type the following into your console and press return to run the code. You will see a lot of red text, but usually that's a good thing!

```{r message=FALSE}
install.packages("tidyverse")
```

TODO: Add figure

Next, we will load the package so it's ready to use:

```{r message=FALSE}
library(tidyverse)
```

TODO: Add figure

::: {.callout-tip}
## What are R Packages?

Packages are collections of R code, data, and documentation that extend the base functionality of R. Think of them like "expansion packs" on top of your basic R software.

Packages are developed by the R community and made available through repositories like CRAN (Comprehensive R Archive Network), Bioconductor, and GitHub. They are especially useful if you want to do a specialized kind of analysis, such as genomic analysis!

We use the `library` command to load and attach packages to the R environment. This means links the package you downloaded to your current session of R.

The "tidyverse" package that you loaded is useful for loading, wrangling, and exploring data.
:::

### Get Differential Expression Data

We'll read in some data from the GEMs website. Run the following code in your console

```{r}
dat <- read_csv("URL where the dataset is stored")
```